EBASE manuscript

### 2023-01-22 manu-draft.qmd
remove mention of Fwoxy in text except for package citation, corrected text about one a value per time step

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.  As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.  The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation called Fwoxy [ForWard OXYgen, @Arriola22].  For Fwoxy, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase. All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.  The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.  An initial oxygen concentration is provided and the simulated time series is created by forward addition of the inputs. As such, the dissolved oxygen output simulated by Fwoxy as change per unit time were modelled by EBASE using the constant physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from Fwoxy:

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with Fwoxy. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).  The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the simulated time series. First, results were compared to the known values after changing characteristics of their prior distributions.  Second, results were compared using a model period of optimization for one day and seven days.  Details of these analyses are as follows.

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).  The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors. For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.  Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).  Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series.  For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for a 24 hour time period if the sampling interval is 15 minutes).  A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.  As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.  An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters.

Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the simulated time series. @fig-apasumdat shows a summary of the comparisons of the simulated data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.  The results are compared using R-squared ($R^2$) values from the linear fit between the simulated data and results from EBASE.  Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.

@@

@@

![Synthetic time series using the Fwoxy model for one year of continuous data at Apalachicola Bay.  Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) simulated $a$ parameter and (c) simulated metabolic estimates for production (P), respiration (R), and gas exchange (D).  The $a$ parameter is shown as an areal value for comparison with the metabolic results.](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a simulated time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the standard deviations are evaluated from low (L), medium (M), to high (H) for a one day and seven day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.  The $b$ parameter is a fixed constant and was not evaluated.](figs/apasumdat.png){#fig-apasumdat}

@@

![Comparison of EBASE results for the best performing model to a simulated time series created with known metabolic parameters. The best performing model for the one day (a, ndays = 1) and seven day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).  The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.  The results are averaged at the time step of the model optimization period.  The $b$ parameter is fixed at 0.251 $\buni$.](figs/optex.png){#fig-optex fig-align="center"}

>>>>

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.  As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.  The default priors were based on those that returned reasonable parameter estimates using a known time series created with forward-feeding oxygen mass balance inputs to @eq-ebase.  [Fwoxy R package, @Arriola22]. All of the physical parameters were held constant, e.g., wind speed, temperature, and salinity.  The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\auni$, 20 $\runi$, and 0.251 $\buni$, respectively.  An initial oxygen concentration was provided as 250 mmol/m$^3$ and the time series was created by forward addition of the inputs. As such, the dissolved oxygen output as change per unit time was modelled by EBASE using the constant the same physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters:

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual physical data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the synthetic data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).  The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series. First, results were compared to the known values after changing characteristics of their prior distributions.  Second, results were compared using a model period of optimization for one day and seven days.  Details of these analyses are as follows.

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).  The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors. For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.  Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).  Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series.  For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model.  A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.  As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.  An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters.

Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series. @fig-apasumdat shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.  The results are compared using R-squared ($R^2$) values from the linear fit between the synthetic data and results from EBASE.  Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.

@@

* Describe how EBASE deals with gaps

@@

![Synthetic time series for one year of continuous data at Apalachicola Bay.  Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for production (P), respiration (R), and gas exchange (D).  The $a$ parameter is shown as an areal value for comparison with the metabolic results.](figs/synapa.png){#fig-synapa}

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the standard deviations are evaluated from low (L), medium (M), to high (H) for a one day and seven day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values.  The $b$ parameter is a fixed constant and was not evaluated.](figs/apasumdat.png){#fig-apasumdat}

@@

![Comparison of EBASE results for the best performing model to a synthetic time series created with known metabolic parameters. The best performing model for the one day (a, ndays = 1) and seven day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).  The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.  The results are averaged at the time step of the model optimization period.  The $b$ parameter is fixed at 0.251 $\buni$.](figs/optex.png){#fig-optex fig-align="center"}
